Measurement Bias from Address Aliasing

### 2014-05-20 paper.tex
Change to ACM TACO formatting
Some adjustments to figures, tables and legends to accomodate
layout.

\documentclass[a4paper,10pt,twocolumn,twoside]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{listings}
\usepackage{authblk}
\usepackage{caption}
\usepackage{mathtools}
\usepackage{underscore}
\usepackage{url}
\usepackage{lmodern}

@@

\author{Lars Kirkholt Melhus}
\author{Rune Erlend Jensen}
\affil{Norwegian University of Science and Technology}
\title{Measurement Bias from Address Aliasing}
\date{} % This is a timeless piece

% Pretty printing of performance counter names
\newcommand{\perfctr}[1] {
  {\lowercase{#1}}
}

@@

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract} % Should be 150 -- 200 words max
    % Motivation -- Problem statement -- Approach -- Results -- Conclusions

    Seemingly innocious properties of the environment, such as the contents of system environment variables, can affect the performance of computer programs.
    This effect is known as \emph{measurement bias}, and has been shown to be significant, but unpredictable and difficult to deal with.

    % In addition to environment variable size, we show how address aliasing can cause measurement bias from variations in dynamically linked heap allocation libraries.
    We find that performance penalties from unfavorable memory address allocations can be more than double the cycle count, even for real world applications.
    Furthermore, the dynamic memory allocation pattern for common libraries like libc tends to give \emph{worst case} performance by default for large allocations.
    An understanding of address aliasing is necessary in order to achieve optimal performance, and we outline how developers can utilize this knowledge.
    \end{abstract}
    \bigskip

  \end{@twocolumnfalse}
]

@@

% This is basically former background section
Bias in performance analysis has been shown to be commonplace and significant in real applications, while being unpredictable and difficult to deal with~\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
Previous work looks at environmental factors such as the size of Unix environment variables, which can affect memory by offsetting the initial position of stack.
Because of the complexity of modern hardware, it is often impossible to predict exactly how things such as environment variables ultimately affect program execution.
Without an accurate understanding of the underlying memory interactions, the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts.
Previous work focuses mostly of how to account for bias by proper sampling, using randomization of execution contexts and statistical methods. % Cite is basically the same as above, all three are about the same -- difficult to split.

@@

% Include preview of aliasing program early to provike interest, also lightly referenced in introduction/background
\begin{lstlisting}[float=t, language=C, caption={Micro-kernel presented by Mytkowicz et al~\cite{Mytkowicz:2009:WrongData}, succeptible to bias from variations to system environment variables.}, label={lst:loopkernel}, frame=lines]
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    for (; g < 65536; g++) {
        i += inc;
        j += inc;
        k += inc; 
    }
    return 0;
}
\end{lstlisting}

@@

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
The number of times this happens can be counted by the following performance counter:
\begin{description}
  \item{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0x07 & 0x01\end{tabular}} \hfill \\
  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.~\cite{Intel:2012:OptimizationManual}
\end{description}

@@

\begin{figure}[float=t]
  \caption{Organization of virtual memory for a 64-bit process.}
  \label{fig:virtualmemory}
  \begin{tikzpicture}[font=\footnotesize]

@@

% Large plot of cycle count for two 4K stack sections.
\pgfplotstableread{bin/micro-kernel-cycles.dat}{\stackoffsettable}
\begin{figure*}[t]
  \caption{Bias from environment size for micro-kernel in Listing~\ref{lst:loopkernel}. Measured average of 10 cycle count samples for 512 different environments. Spikes show aliasing case, occurring once for each 4K period}
  \label{fig:envbias}
  \begin{tikzpicture}
    \begin{axis}[
        width = \textwidth,
        height = 6cm,
        font = \small,
        xlabel=Bytes added to environment,
        ylabel=Cycles,
        domain = 0:8192,
        xtick = {0,1024,...,8192},
        xmin = 0,
        xmax = 8192,
        cycle list name = exotic
      ]
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

% Detailed comparison between median and worst cases.
\begin{table}[t]
  \caption{Events with significant correlation to cycle count.}
  \label{tab:loopcorrelation}
  %\centering
  \renewcommand{\tabcolsep}{1pt} % Less padding between cells
  \pgfplotstabletypeset[
    font=\scriptsize,
    int detect, % Output whole numbers for counter values
    col sep=comma,
    columns={Performance counter, Median, [index]3, [index]4},
    column type=r,
    columns/Performance counter/.style={
      string type, 
      column type=l,
      postproc cell content/.append code={
        \pgfkeysalso{@cell content=\perfctr{##1}}
      }
    },
    every head row/.style={
      output empty row,
      before row={\toprule
        Performance counter & Median & Spike 1 & Spike 2 \\
        },
      after row=\midrule,
    },
    every last row/.style={after row=\bottomrule}
  ]{bin/micro-kernel-comparison.csv}
\end{table}

@@

% See analysis for results showing this does not alias
\begin{lstlisting}[float=t, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
#define ALIAS(a, b) \
    (((long)&a) & 0xfff == ((long)&b) & 0xfff)
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    if (ALIAS(inc, i) || ALIAS(g, i))
        return main();
    for (; g < 65536; g++) {
        i += inc;
        j += inc;
        k += inc;
    }
    return 0;
}
\end{lstlisting}

@@

\pgfplotstableread{bin/conv-default-o2.estimate.dat}{\convtabletwo}
\pgfplotstableread{bin/conv-default-o3.estimate.dat}{\convtablethree}
\begin{figure*}[t]
  \caption{Estimated cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Listing~\ref{lst:conv}. Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Input size $n=2^{20}$.}
  \label{fig:conv-default}
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O2},
        font = \small,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwo ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwo ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O3},
        font = \small,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethree ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethree ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

% Table with detailed performance data, use estimated for O2 because it does not need as many columns (!)
\begin{table*}[t]
  \caption{Most interesting performance counters and correlation with cycle count for optimization O2, estimated cost accounting for constant overhead.}
  \label{tab:convstats}
  \pgfplotstabletypeset[
    font=\small,
    int detect, % Output whole numbers for counter values
    col sep=comma,
    columns={Performance counter, Correlation, 0, 2, 4, 8},
    column type=r,
    columns/Performance counter/.style={
      string type, 
      column type=l,
      postproc cell content/.append code={
        \pgfkeysalso{@cell content=\perfctr{##1}}
      }
    },
    columns/Correlation/.style={
      fixed,
      fixed zerofill,
      precision=2
    },
    every head row/.style={
      output empty row,
      before row={\toprule
        Performance counter & $r$ & 0 & 2 & 4 & 8 \\
        },
      after row=\midrule,
    },
    every last row/.style={after row=\bottomrule}
  ]{bin/conv-default-o2.estimate.csv}
\end{table*}

@@

\begin{table*}[t]
  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}
  \label{tab:mallocompare}
  \pgfplotstabletypeset[
    font=\small,
    col sep=comma,
    string type,
    columns={Allocation, 64, 5120, 65536, 1048576, 4194304},
    column type=r,
    columns/Allocation/.style={
      string type, 
      column type=l
    },
    every head row/.style={
      output empty row,
      before row={\toprule
         & 64 B & 5,120 B & 65,536 B & 1,048,576 B & 4,194,304 B \\
        },
      after row=\midrule,
    },
    every odd row/.style={after row=\midrule},
    every last row/.style={after row=\bottomrule}
  ]{bin/malloc-comparison.csv}
\end{table*}

>>>>

% Typeset for Transactions on Architecture and Code Optimization
\documentclass[prodmode,acmtaco]{acmsmall}
%\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{booktabs}
%\usepackage{tabularx}
\usepackage{tikz}
\usepackage{listings}
%\usepackage{authblk} % acm error, \affil
\usepackage{caption} % 
\usepackage{mathtools}
\usepackage{underscore}
\usepackage{url}
\usepackage{lmodern} % Much clearer text

@@

\newcommand{\perfctr}[1] {
  {\lowercase{#1}}
}

%\acmVolume{V}
%\acmNumber{N}
%\acmArticle{A}
%\acmYear{YYYY}
%\acmMonth{0}

% Page headers
\markboth{L. K. Melhus and R. E. Jensen}{Measurement Bias from Address Aliasing}

\title{Measurement Bias from Address Aliasing}
\author{
LARS KIRKHOLT MELHUS and RUNE ERLEND JENSEN
  \affil{Norwegian University of Science and Technology}}

% Motivation -- Problem statement -- Approach -- Results -- Conclusions
\begin{abstract} % Should be 150 -- 200 words max
Seemingly innocious properties of the environment, such as the contents of system environment variables, can affect the performance of computer programs.
This effect is known as \emph{measurement bias}, and has been shown to be significant, but unpredictable and difficult to deal with.

% In addition to environment variable size, we show how address aliasing can cause measurement bias from variations in dynamically linked heap allocation libraries.
We find that performance penalties from unfavorable memory address allocations can be more than double the cycle count, even for real world applications.
Furthermore, the dynamic memory allocation pattern for common libraries like libc tends to give \emph{worst case} performance by default for large allocations.
An understanding of address aliasing is necessary in order to achieve optimal performance, and we outline how developers can utilize this knowledge.
\end{abstract}

%\category{}{}{}
\terms{Measurement bias, Optimization}
\keywords{4K address aliasing, Intel ``Haswell'' architecture, heap allocators}
\acmformat{Lars K. Melhus and Rune E. Jensen, 2014. Measurement Bias from Address Aliasing}

@@

\begin{bottomstuff}

\end{bottomstuff}

\maketitle

@@

% This is basically former background section
Bias in performance analysis has been shown to be commonplace and significant in real applications, while being unpredictable and difficult to deal with~%\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
Previous work looks at environmental factors such as the size of Unix environment variables, which can affect memory by offsetting the initial position of stack.
Because of the complexity of modern hardware, it is often impossible to predict exactly how things such as environment variables ultimately affect program execution.
Without an accurate understanding of the underlying memory interactions, the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts.
Previous work focuses mostly of how to account for bias by proper sampling, using randomization of execution contexts and statistical methods. % Cite is basically the same as above, all three are about the same -- difficult to split.

@@

% Include preview of aliasing program early to provike interest, also lightly referenced in introduction/background
% Cite does not work in acm style
% Micro-kernel presented by Mytkowicz et al~\cite{Mytkowicz:2009:WrongData}, 
\begin{lstlisting}[float=t, language=C, caption={Micro-kernel succeptible to bias from variations to system environment variables.}, label={lst:loopkernel}, frame=lines]
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    for (; g < 65536; g++) {
        i += inc;
        j += inc;
        k += inc; 
    }
    return 0;
}
\end{lstlisting}

@@

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
The number of times this happens can be counted by the following performance counter:
\begin{description}
  \item LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0x07 & 0x01\end{tabular}} \hfill \\
  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.~\cite{Intel:2012:OptimizationManual}
\end{description}

@@

\begin{figure}[h]
  \centering
  \caption{Organization of virtual memory for a 64-bit process.}
  \label{fig:virtualmemory}
  \begin{tikzpicture}[font=\footnotesize]

@@

% Large plot of cycle count for two 4K stack sections.
\pgfplotstableread{bin/micro-kernel-cycles.dat}{\stackoffsettable}
\begin{figure*}[t]
  \caption{Bias from environment size for micro-kernel in Listing~\ref{lst:loopkernel}. Measured average of 10 cycle count samples for 512 different environments. Spikes show aliasing case, occurring once for each 4K period}
  \label{fig:envbias}
  \begin{tikzpicture}
    \begin{axis}[
        width = \textwidth,
        height = 6cm,
        font = \footnotesize,
        xlabel=Bytes added to environment,
        ylabel=Cycles,
        domain = 0:8192,
        xtick = {0,1024,...,8192},
        xmin = 0,
        xmax = 8192,
        cycle list name = exotic
      ]
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

% Detailed comparison between median and worst cases.
\begin{table}[b]
  \centering
  \caption{Events with significant correlation to cycle count}
  \label{tab:loopcorrelation}
  %\renewcommand{\tabcolsep}{1pt} % Less padding between cells
  \renewcommand{\tabcolsep}{4pt} % Less padding between cells
  \pgfplotstabletypeset[
    font=\footnotesize,
    int detect, % Output whole numbers for counter values
    col sep=comma,
    columns={Performance counter, Median, [index]3, [index]4},
    column type=r,
    columns/Performance counter/.style={
      string type, 
      column type=l,
      postproc cell content/.append code={
        \pgfkeysalso{@cell content=\perfctr{##1}}
      }
    },
    every head row/.style={
      output empty row,
      before row={\toprule
        Performance counter & Median & Spike 1 & Spike 2 \\
        },
      after row=\midrule,
    },
    every last row/.style={after row=\bottomrule}
  ]{bin/micro-kernel-comparison.csv}
\end{table}

@@

% See analysis for results showing this does not alias
\begin{lstlisting}[float=b, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
#define ALIAS(a, b) \
    (((long)&a) & 0xfff == ((long)&b) & 0xfff)
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    if (ALIAS(inc, i) || ALIAS(g, i))
        return main();
    for (; g < 65536; g++) {
        i += inc;
        j += inc;
        k += inc;
    }
    return 0;
}
\end{lstlisting}

@@

\pgfplotstableread{bin/conv-default-o2.estimate.dat}{\convtabletwo}
\pgfplotstableread{bin/conv-default-o3.estimate.dat}{\convtablethree}
\begin{figure*}[t]
  \centering
  \caption{Estimated cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Listing~\ref{lst:conv}. Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Input size $n=2^{20}$.}
  \label{fig:conv-default}
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O2},
        font=\footnotesize,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white,
        width=\textwidth/1.9, % Make it fit to text width side by side
        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwo ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwo ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O3},
        font=\footnotesize,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white,
        width=\textwidth/1.9,
        skip coords between index={20}{32}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethree ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethree ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

% Table with detailed performance data, use estimated for O2 because it does not need as many columns (!)
\begin{table*}[t]
  \centering
  \caption{Most interesting performance counters and correlation with cycle count for optimization O2, estimated cost accounting for constant overhead.}
  \label{tab:convstats}
  \renewcommand{\tabcolsep}{4pt} % Less padding between cells
  \pgfplotstabletypeset[
    font=\footnotesize,
    int detect, % Output whole numbers for counter values
    col sep=comma,
    columns={Performance counter, Correlation, 0, 2, 4, 8},
    column type=r,
    columns/Performance counter/.style={
      string type, 
      column type=l,
      postproc cell content/.append code={
        \pgfkeysalso{@cell content=\perfctr{##1}}
      }
    },
    columns/Correlation/.style={
      fixed,
      fixed zerofill,
      precision=2
    },
    every head row/.style={
      output empty row,
      before row={\toprule
        Performance counter & $r$ & 0 & 2 & 4 & 8 \\
        },
      after row=\midrule,
    },
    every last row/.style={after row=\bottomrule}
  ]{bin/conv-default-o2.estimate.csv}
\end{table*}

@@

\begin{table*}[t]
  \centering
  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}
  \label{tab:mallocompare}
  \renewcommand{\tabcolsep}{4pt} % Less padding between cells
  \pgfplotstabletypeset[
    font=\footnotesize,
    col sep=comma,
    string type,
    columns={Allocation, 64, 5120, 65536, 1048576, 4194304},
    column type=r,
    columns/Allocation/.style={
      string type, 
      column type=l
    },
    every head row/.style={
      output empty row,
      before row={\toprule
         & 64 B & 5,120 B & 65,536 B & 1,048,576 B & 4,194,304 B \\
        },
      after row=\midrule,
    },
    every odd row/.style={after row=\midrule},
    every last row/.style={after row=\bottomrule}
  ]{bin/malloc-comparison.csv}
\end{table*}
